### 2012-10-24 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
spell check for prelim

In the context of solving standalone linear systems, Monte Carlo
methods do not provide significant merit over Krylov methods due to
the fact that the linear operator must be explicitly formed. For many
applications, such a requirement is prohibitive and perhaps not even
feasible to implement. Therefore, a Monte Carlo solver is best suited
for situations in which not only are Krylov methods applicable, but
also in which the operator is readily, if not naturally,
formed. Modern nonlinear methods meet both of these requirements with
Newton methods used in conjuction with Krylov methods for a robust,
fully implicit solution strategy. Furthermore, modern techniques exist
that permit the automatic construction of the linear operator
generated within a Newton method based on the nonlinear residual
evaluations, providing all of the components necessary for a Monte
Carlo solver to provide value. We therefore devise a new nonlinear
method based on the MCSA algorithm and Newton's method and discuss its
potential benefits.

>>>>

In the context of solving standalone linear systems, Monte Carlo
methods do not provide significant merit over Krylov methods due to
the fact that the linear operator must be explicitly formed. For many
applications, such a requirement is prohibitive and perhaps not even
feasible to implement. Therefore, a Monte Carlo solver is best suited
for situations in which not only are Krylov methods applicable, but
also in which the operator is readily, if not naturally,
formed. Modern nonlinear methods meet both of these requirements with
Newton methods used in conjunction with Krylov methods for a robust,
fully implicit solution strategy. Furthermore, modern techniques exist
that permit the automatic construction of the linear operator
generated within a Newton method based on the nonlinear residual
evaluations, providing all of the components necessary for a Monte
Carlo solver to provide value. We therefore devise a new nonlinear
method based on the MCSA algorithm and Newton's method and discuss its
potential benefits.

### 2012-10-24 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
spell check for prelim

\section{Adjoint Method}
\label{sec:adjoint_mc}
An alternative formulation for Monte Carlo matrix inversion is the
adjoint method. We begin by defining the linear system adjoint to
Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator. We can split this
equation to mirror Eq~(\ref{eq:richardson_split}) by defining the
following inner product equivalence \citep{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
With this statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
  \label{eq:adjoint_split_system}
\end{equation}
As was required for convergence with the direct method using
Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius. From this definition it
follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Using these definitions, we can derive an estimator from the adjoint
method that will also give the solution vector, $\ve{x}$. As with the
direct method, we can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the direct solution. We
achieve this by using Eq~(\ref{eq:adjoint_vector_relation}) as a
constraint. Here we have 2 unkowns, $\ve{y}$ and $\ve{d}$ and
therefore we require a second constraint to close the system. As a
second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the direct source and
the adjoint solution using a delta function source, we recover the
direct solution we are looking for

@@

\section{Sequential Monte Carlo}
\label{sec:sequential_mc}
The direct and adjoint Neumann-Ulam methods described are limited by a
convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
$N$ is the number of random walk permutations. In 1962, Halton
presented a residual Monte Carlo method that moves towards exponential
convergence rates \citep{halton_sequential_1962} and further refined
his work some years later \citep{halton_sequential_1994}. Applications
of his work by the transport community have confirmed convergence
rates on the order of of $exp(-N)$ \citep{evans_residual_2003}. In
much the same way as projection methods, Halton's method, sequential
Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
directly reducing the residual vector. He proposed the following
iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}$ is computed by the adjoint
Monte Carlo method. The merits of Halton's approach are immediately
visible in that we have now broken the binding of the convergence rate
to the Central Limit Thereom. Here, the Monte Carlo solver is used to
produce a correction from the residual, analogous to using the
residual to extract a correction from the search subspace in a
projection method. By doing this, the Monte Carlo error is bound in
the correction used to update the solution and therefore does not
explicitly manifest itself in the overall convergence of the
solution. The downside of such a method is that if the solution guess
is poor, then many iterations are required in order to reach
exponential converge as the Monte Carlo error (and therefore the
Central Limit Theorem) does dominate in this situation. Therefore, if
a good initial guess is not available, Halton's method is still
characterized by poor performance.

@@

\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
\label{subsubsec:msod}
In 2010, Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed the similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1,000)$ times achieves simulation on $O(100,000)$
processors, thus providing spatial and particle parallelism. Each
collection of processors that constitutes a representation of the
entire domain is referred to as a set, and within a set overlap occurs
among its subdomains. The original motivation was to decompose the
domain in a way that it remained in a physical cabinet in a large
distributed machine, thus reducing latency costs during
communication. A multiple set scheme is also motivated by the fact
that communication during particle transport only occurs within a set,
limiting communications during the transport procedure to a group of
$O(100)$ processors, a number that was shown to have excellent
parallel efficiencies in Brunner's work and therefore will scale well
in this algorithm. The overlapping domains within each set also
demonstrated reduced communication costs. On each processor, the
source is sampled in the local domain that would exist if no overlap
was used while tallies can be made over the entire overlapping domain.

@@

What they found was that there was indeed an optimum ratio of local
domain to global domain size that results in a minimum total compute
time. They also found that this ratio was only dependent on the
latency and bandwidth constraints of the parallel machine. We expect
this because nearest neighbor communication patterns scale with the
volume and surface sizes of the local domains. The larger the volume,
the more compute time is spent within the local volume and therefore
at some point you are limited by the latency of the system, or the
time it takes to receive the data to operate on. However, at a certain
point, when the surface of those volumes becomes large enough, the
amount of data that must be communicated grows large enough that it
may overcome the bandwidth limitations of the system, resulting in a
decrease in performance. In their numerical experiments, they used the
above geometry and leakage rates based on that computed for a reactor
system for a single material problem with a total cross section
equivalent to that of homogonized light water reactor. Given the
optimal local domain size, they observed for load balanced situations
that nearest neighbor communication costs during transport are a small
fraction of total compute time, leading to the conclusing that domain
decomposition is feasible for large-scale Monte Carlo calculations.

\subsubsection{Load Balancing Concerns}
\label{subsubsec:mc_load_balancing}
Although domain decomposition was shown to be efficient in a prefectly
load balanced situation in Siegel's work \citep{siegel_analysis_2012},
careful consideration must be made for situations where this is not
the case. Given the stochastic nature of the problem and lack of a
globally homogeneous domain, parallel Monte Carlo simulations are
inherently load imbalanced. Procassini and others worked to alleviate
some of the load imbalances that are generated by both particle and
spatial parallelism and are therefore applicable to the MSOD algorithm
\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
in which the number of times a particular domain was replicated was
dependent on the amount of work in that domain (i.e. domains with a
high particle flux and therefore more particle histories to compute
require more work). In this variation, domains that require more work
will be replicated more frequently at reduced particle counts in each
replication. Furthermore, Procassini and colleagues noted that as the
simulation progressed and particles were transported throughout the
domain, the amount of replication for each domain would vary as
particle histories began to diffuse, causing some regions to have
higher work loads and some to have smaller work loads than the initial
conditions.

>>>>

\section{Adjoint Method}
\label{sec:adjoint_mc}
An alternative formulation for Monte Carlo matrix inversion is the
adjoint method. We begin by defining the linear system adjoint to
Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator. We can split this
equation to mirror Eq~(\ref{eq:richardson_split}) by defining the
following inner product equivalence \citep{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
With this statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
  \label{eq:adjoint_split_system}
\end{equation}
As was required for convergence with the direct method using
Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius. From this definition it
follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Using these definitions, we can derive an estimator from the adjoint
method that will also give the solution vector, $\ve{x}$. As with the
direct method, we can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the direct solution. We
achieve this by using Eq~(\ref{eq:adjoint_vector_relation}) as a
constraint. Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and
therefore we require a second constraint to close the system. As a
second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the direct source and
the adjoint solution using a delta function source, we recover the
direct solution we are looking for

@@

\section{Sequential Monte Carlo}
\label{sec:sequential_mc}
The direct and adjoint Neumann-Ulam methods described are limited by a
convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
$N$ is the number of random walk permutations. In 1962, Halton
presented a residual Monte Carlo method that moves towards exponential
convergence rates \citep{halton_sequential_1962} and further refined
his work some years later \citep{halton_sequential_1994}. Applications
of his work by the transport community have confirmed convergence
rates on the order of of $exp(-N)$ \citep{evans_residual_2003}. In
much the same way as projection methods, Halton's method, sequential
Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
directly reducing the residual vector. He proposed the following
iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}$ is computed by the adjoint
Monte Carlo method. The merits of Halton's approach are immediately
visible in that we have now broken the binding of the convergence rate
to the Central Limit Theorem. Here, the Monte Carlo solver is used to
produce a correction from the residual, analogous to using the
residual to extract a correction from the search subspace in a
projection method. By doing this, the Monte Carlo error is bound in
the correction used to update the solution and therefore does not
explicitly manifest itself in the overall convergence of the
solution. The downside of such a method is that if the solution guess
is poor, then many iterations are required in order to reach
exponential converge as the Monte Carlo error (and therefore the
Central Limit Theorem) does dominate in this situation. Therefore, if
a good initial guess is not available, Halton's method is still
characterized by poor performance.

@@

\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
\label{subsubsec:msod}
In 2010, Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed the similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1,000)$ times achieves simulation on $O(100,000)$
processors, thus providing spatial and particle parallelism. Each
collection of processors that constitutes a representation of the
entire domain is referred to as a set, and within a set overlap occurs
among its sub-domains. The original motivation was to decompose the
domain in a way that it remained in a physical cabinet in a large
distributed machine, thus reducing latency costs during
communication. A multiple set scheme is also motivated by the fact
that communication during particle transport only occurs within a set,
limiting communications during the transport procedure to a group of
$O(100)$ processors, a number that was shown to have excellent
parallel efficiencies in Brunner's work and therefore will scale well
in this algorithm. The overlapping domains within each set also
demonstrated reduced communication costs. On each processor, the
source is sampled in the local domain that would exist if no overlap
was used while tallies can be made over the entire overlapping domain.

@@

What they found was that there was indeed an optimum ratio of local
domain to global domain size that results in a minimum total compute
time. They also found that this ratio was only dependent on the
latency and bandwidth constraints of the parallel machine. We expect
this because nearest neighbor communication patterns scale with the
volume and surface sizes of the local domains. The larger the volume,
the more compute time is spent within the local volume and therefore
at some point you are limited by the latency of the system, or the
time it takes to receive the data to operate on. However, at a certain
point, when the surface of those volumes becomes large enough, the
amount of data that must be communicated grows large enough that it
may overcome the bandwidth limitations of the system, resulting in a
decrease in performance. In their numerical experiments, they used the
above geometry and leakage rates based on that computed for a reactor
system for a single material problem with a total cross section
equivalent to that of homogenized light water reactor. Given the
optimal local domain size, they observed for load balanced situations
that nearest neighbor communication costs during transport are a small
fraction of total compute time, leading to the conclusion that domain
decomposition is feasible for large-scale Monte Carlo calculations.

\subsubsection{Load Balancing Concerns}
\label{subsubsec:mc_load_balancing}
Although domain decomposition was shown to be efficient in a perfectly
load balanced situation in Siegel's work \citep{siegel_analysis_2012},
careful consideration must be made for situations where this is not
the case. Given the stochastic nature of the problem and lack of a
globally homogeneous domain, parallel Monte Carlo simulations are
inherently load imbalanced. Procassini and others worked to alleviate
some of the load imbalances that are generated by both particle and
spatial parallelism and are therefore applicable to the MSOD algorithm
\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
in which the number of times a particular domain was replicated was
dependent on the amount of work in that domain (i.e. domains with a
high particle flux and therefore more particle histories to compute
require more work). In this variation, domains that require more work
will be replicated more frequently at reduced particle counts in each
replication. Furthermore, Procassini and colleagues noted that as the
simulation progressed and particles were transported throughout the
domain, the amount of replication for each domain would vary as
particle histories began to diffuse, causing some regions to have
higher work loads and some to have smaller work loads than the initial
conditions.
